Bryan Flores
ADS-503
Final Project: Exploratory Analysis

```{r}
setwd("C:/Users/Rudy/OneDrive/MADS/503_Applied Predictive Modeling/finalproject/data/")
# CSV Upload
met_21_22 = read.csv("met_21_22_9755371.csv", header=TRUE, stringsAsFactors=FALSE)
met_22_23 = read.csv('met_22_23_9755371.csv', header=TRUE, stringsAsFactors = FALSE)
wl_21_23 = read.csv('wl_21_23_9755371.csv', header=TRUE, stringsAsFactors = FALSE)
pacIOOS_21_22 = read.csv('dhw_5km_6a81_de46_d4a4.csv', header=TRUE, stringsAsFactors = FALSE)
pacIOOS_22_23 = read.csv('dhw_5km_db8d_5daf_80ed.csv', header=TRUE, stringsAsFactors = FALSE)
```
```{r}
# Examine the input table structures
str(met_21_22)
str(met_22_23)
str(wl_21_23)
str(pacIOOS_21_22)
str(pacIOOS_22_23)
```

```{r}
# Combine years for meteorological & coral tables
library('dplyr')
met21_23 = bind_rows(met_21_22, met_22_23)
coral21_23 = bind_rows(pacIOOS_21_22, pacIOOS_22_23)
```
```{r}
# Examine the meteorological tables
dim(met21_23)
colnames(met21_23)
str(met21_23)
```
```{r}
# Examine the water level table
dim(wl_21_23)
colnames(wl_21_23)
str(wl_21_23)
```
```{r}
# Examine the coral data
dim(coral21_23)
colnames(coral21_23)
str(coral21_23)
```
```{r}
# Convert coral table column names to lowercase
names(coral21_23) = tolower(names(coral21_23))
coral21_23
```
```{r}
# Convert coral datetime column to a standard date format. YYYY-MM-DD
## The coral timestamp field marks 6 data points at the beginning of each day.
### Extracting date from timestamp column
coral21_23$date = format(as.POSIXct(coral21_23$time), format="%Y-%m-%d")
```

There are now 3 tables each span 2021 through 2023 with a standard date format.

```{r}
# Count total missing values in each column of data frame
## Coral
sapply(coral21_23, function(x) sum(is.na(x)))
## Meteorological
sapply(met21_23, function(x) sum(is.na(x)))
## Water Level
sapply(wl_21_23, function(x) sum(is.na(x)))
```

There are no missing values in any of the tables. 
---
All Coral _mask suffix columns contain only zeros and will be dropped from the data frame since we cannot confirm whether this an accurate or a missed reading based on the information provided from its source. The date column was created by removing the timestamp from the time column.

Coral columns to be dropped: time, crw_baa_mask, crw_dhw_mask, crw_hotspot_mask, and crw_sst_anomaly_mask
---
The meteorological data contains 5 columns with not null empty values and shall be dropped. Time will also be dropped due to an inaccuracy in the mapping with the coral table.

Meteorological columns to be dropped: time, wind_speed, wind_direction, wind_gust, humidity, and visibility
---
The Water Level (WL) inf column contains no information and shall be dropped.

```{r}
# Dropping empty columns from each table
coral21_23_sub = subset(coral21_23, select = -c(crw_baa_mask, crw_dhw_mask, crw_hotspot_mask,
                                                crw_sstanomaly_mask, time))
met21_23_sub = subset(met21_23, select = -c(wind_speed, wind_direction, wind_gust, humidity,
                                            visibility, time))
wl_21_23_sub = subset(wl_21_23, select = -c(inf, time))
```

The coral table has 6 measurements per day with a timestamp of 00:00, which is why it was removed. The meteorological table has 1 measurement per hour per day. Lastly, the water level table has 1 measurement per month. 

Coral: This is the central table the meteorological and water level tables will be manipulated for.

Meteor.: The ideal mapping would be to break up the hourly measurements into 6 4-measurement groups. The caveat to this approach in our case is that there is no way to be which order the coral measurements were taken -- simultaneously or sequentially. Therefore, we will use the daily average of air temperature and barometric metric pressure and apply a one-to-many mapping with the coral measurements. Air temperature and barometric pressure will be represented 6 times per day.

Water Level: Apply a one-to-many mapping with the coral table. Each monthly water level measurement will be represented 6 times per day per month. Water level measurements were only taken on the first of every month. Therefore, we will have to join on the yyyy-mm of the date column. For simplicity, I will create a new year-month column in each dataframe, then drop it once the merge is complete.

```{r}
# Convert air temp & baro columns to numeric
met21_23_sub = transform(met21_23_sub, air_temp = as.numeric(air_temp), baro = as.numeric(baro))
```
```{r}
# Find daily averages: air temperature & barometric pressure
met21_23_sub_avg = aggregate(met21_23_sub[, 2:3], list(met21_23_sub$date), mean)
met21_23_sub_avg
```
```{r}
# Change new subset column name
colnames(met21_23_sub_avg) = c('date', 'air_temp', 'baro')
# Convert air temperature to celcius from fahrenheit
library(weathermetrics)
met21_23_sub_avg$air_temp = fahrenheit.to.celsius(met21_23_sub_avg$air_temp, round = 2)
```

```{r}
# Map Meteorological to Coral
coral_met = merge(x = coral21_23_sub, y = met21_23_sub_avg, by="date", all.x=TRUE)
```
```{r}
# Map Water Level to Coral/Met Table
## Add year-month columns for merging
coral_met$ym = format(as.Date(coral_met$date), "%Y-%m")
wl_21_23_sub$ym = format(as.Date(wl_21_23_sub$date), "%Y-%m")

cmw_merged = merge(x = coral_met, y = wl_21_23_sub, by="ym", all.x=TRUE)
## Dropping year-month column
cmw_merged = subset(select(cmw_merged, -c(ym, latitude, longitude, date.y)))
## Rename date column
colnames(cmw_merged)[1] = "date"
```
```{r}
# Storing Final Set as csv
setwd("C:/Users/Rudy/Desktop/DataScience/deepseacorals/data/")
write.csv(cmw_merged, "coral_pretrain.csv")
```


The final set before exploratory analysis and feature selection is cmw_merged. The year-month column, ym, has been removed along with latitude and longitude. Latitude and longitude were necessary in obtaining the data, however, the area of the measurements taken will remain constant over the period of our data set. 

Our final set contains 4386 rows and 19 columns -- 6 measurements per day per month from 5/27/2021 through 5/27/2023. CRW_BAA is the predictor column is the bleaching alert area containing 5 classes: 0 for no stress, 1 for bleaching watch, 2 for bleaching warning, 3 for bleaching alert level 1, and 4 for bleaching alert level 2. 

--- 
EXPLORATORY ANALYSIS
```{r}
# function to count number of missing rows
countmissing <- function(x) {
sum(is.na(x))
}
# summarize missing data by predictor
apply(cmw_merged, 2, countmissing)
```
There are 348 missing values for each of the water level predictors and 192 missing values for air temperature and barometric pressure. The readings were taken sequentially as our data is a time series data set; the imputeTS library is used to impute each predictors missing values with the moving average. The imputeTS package specializes on (univariate) time series imputation. The moving average was chosen for imputation because water levels in Puerto Rico change with the tides. Puerto Rico experiences consistent but nontrivial tide changes. Peak water levels per month are just under 3 ft, whereas the trough is approximately -0.5 ft. 

The imputeTS function, na_ma, imputes missing values by the weighted moving average. It uses a semi-adaptive window size that ensures all missing values are replaced. In this case, the default exponential moving average (EWMA), which is a recursive function that calculates the current observation with the previous one. EWMA uses weighting factors that decrease exponentially. The na_ma function accepts an integer width of the moving average e.g. k=3, which means the sum of 3 observations to the left and 3 observations to the right of the NA value are taken into account. Additionally, if those 6 total values are also NA, the window is automatically increased until there are at least 3 non-NA values present. We use the default value of k=4. 

```{r}
# Impute missing values w/ the moving average
library(imputeTS)
cmw_merged$air_temp = na_ma(cmw_merged$air_temp)
cmw_merged$baro = na_ma(cmw_merged$baro)
cmw_merged$highest = na_ma(cmw_merged$highest)
cmw_merged$mhhw = na_ma(cmw_merged$mhhw)
cmw_merged$mhw = na_ma(cmw_merged$mhw)
cmw_merged$msl = na_ma(cmw_merged$msl)
cmw_merged$mtl = na_ma(cmw_merged$mtl)
cmw_merged$mlw = na_ma(cmw_merged$mlw)
cmw_merged$mllw = na_ma(cmw_merged$mllw)
cmw_merged$lowest = na_ma(cmw_merged$lowest)
```
```{r}
countmissing <- function(x) {
sum(is.na(x))
}
# Re-count missing values
apply(cmw_merged, 2, countmissing)
```
There are no more missing values in the pre-training set.

```{r}
# library(tidyverse)
library(ggplot2)
# Distribution of the bleaching hotspot
hist(cmw_merged$crw_hotspot, xlab="Hotspot (Celsius)", main = "Frequency Histogram of Bleaching Hotspot", col = "seagreen3")
```
```{r}
# Distribution of bleaching alert area
hist(cmw_merged$crw_baa, xlab="Alert Area", main = "Frequency Histogram of Bleaching Alert Area", col = "seagreen3")
```

```{r}
# Difference in hotspot temperatures between bleaching alert area classes
boxplot(crw_hotspot ~ crw_baa, data=cmw_merged, col="salmon", 
        xlab = "Alert Area", ylab = "Hotspot", main = "Coral Bleaching Hotspot & Alert Area")
```

```{r}
# Distribution of predictors
ggplot(data=cmw_merged, aes(x=crw_hotspot, group=crw_baa, fill=crw_baa)) +
    geom_density(adjust=1.5, alpha=.6)
```
```{r}
# Load required library
library(plotly)
 
# Set discrete predictors as variables
x1 = cmw_merged$crw_baa
x2 = cmw_merged$crw_baa_7d_max
x3 = cmw_merged$crw_baa_7d_max_mask

# Create Plotly figure with multiple histograms
plot_ly() %>%
  add_histogram(x = ~x1, name = "Alert Area", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x2, name = "7-Day Max", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x3, name = "7-Day Max Mask", nbinsx = 30, opacity = 0.7) %>%
  layout(title = "Freq. Distribution of Discrete Predictors",
         xaxis = list(title = "X Values"),
         yaxis = list(title = "Frequency"))
```

Interpretation of discrete predictors: Based on the above plots, there are slightly normal distributions for Alert Area (crw_baa) and 7-Day Max (crw_baa_7d_max). Alternatively, 7-Day Max Mask (crw_baa_7d_max_mask) has a skewed distribution. All of its values cluster at 0. Thus, 7-Day Max Mask will be removed from the dataset.

```{r}
# Removing 7-Day Max Mask
cmw_merged = subset(select(cmw_merged, -c(crw_baa_7d_max_mask)))
```

```{r}
# Load required library
library(plotly)
 
# Set continous predictors as variables
x1 = cmw_merged$crw_dhw
x2 = cmw_merged$crw_hotspot
x3 = cmw_merged$crw_sst
x4 = cmw_merged$crw_sstanomaly
x5 = cmw_merged$air_temp

# Create Plotly figure with multiple histograms
plot_ly() %>%
  add_histogram(x = ~x1, name = "Degree Heat Week", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x2, name = "Hotspot", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x3, name = "Sea Surface Temp", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x4, name = "SST Anomaly", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x5, name = "Air Temp", nbinsx = 30, opacity = 0.7) %>%
  layout(title = "Freq. Distribution of Coral Reef Predictors",
         xaxis = list(title = "X Values"),
         yaxis = list(title = "Frequency"))
```
Interpretation of CRW Continuous variables:
- Degree heat week (crw_dhw), blue, has a skewed distribution. DHW values range between 0 and 100 C. Over the past two years, the measurements cluster heavily at zero and a small portion around the value 2.
- Hotspot (crw_hotspot), orange, has an approximately normal distribution.
- SST Anomaly (crw_sstanomaly), red, has a skewed distribution. Though SST Anomaly has a limited range of -15 to 15 degrees Celcius, the majority of the data is centered at 0.5 Celcius. 
- Sea Surface Temp (crw_sst, green) and Air Temperature (air_temp, purple) have approximately normal distributions.

```{r}
# Load required library
library(plotly)
 
# Set continous predictors as variables
x1 = cmw_merged$highest
x2 = cmw_merged$mhhw 
x3 = cmw_merged$mhw 
x4 = cmw_merged$msl 
x5 = cmw_merged$mtl 
x6 = cmw_merged$mlw 
x7 = cmw_merged$mllw 
x8 = cmw_merged$lowest 

# Create Plotly figure with multiple histograms
plot_ly() %>%
  add_histogram(x = ~x1, name = "Highest", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x2, name = "Mean Higher High Water", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x3, name = "Mean High Water", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x4, name = "Mean Sea Level", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x5, name = "Mean Tide Level", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x3, name = "Mean Low Water", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x4, name = "Mean Lower Low Water", nbinsx = 30, opacity = 0.7) %>%
  add_histogram(x = ~x5, name = "Trough Water Level", nbinsx = 30, opacity = 0.7) %>%
  layout(title = "Freq. Distribution of Water Level Predictors",
         xaxis = list(title = "X Values"),
         yaxis = list(title = "Frequency"))
```

All water level variables have approximate normal distributions.

```{r}
# Boxplots of Coral Reef Watch predictors
boxplot(subset(select(cmw_merged, c(crw_dhw, crw_hotspot, crw_sst, crw_sstanomaly, air_temp))),
        main="Coral Reef Watch Boxplots",
        xlab="Predictors",
        ylab="Temp (C)")
```
Based on the above boxplots, all variables except crw_dhw have a symmetric distribution. The crw_dhw shows a median line at the bottom of the box with no visible lower whisker meaning the distribution is positively skewed. Additionally, hotspot, sst, and air_temp have a moderate dispersion whereas sstanomaly and dhw are tightly compact.

```{r}
# Boxplot of barometer readings
boxplot(cmw_merged$baro, main="Barometric Pressure Boxplot", xlab="Barometric Pressure", ylab="Pressure (mb)")
```
Barometric pressure (baro) has a symmetric distribution with outliers below the lower quartile. 

```{r}
# Boxplots of meteorological & water level predictors
boxplot(subset(select(cmw_merged, c(highest, mhhw, mhw, msl, mtl, mlw, mllw, lowest))),
        main="Water Level Boxplots",
        xlab="Predictors",
        ylab="Level (ft)")
```

All predictors have approximately the same dispersion, which is moderate. However, highest has a dispersion that is almost twice as large as the others. All median lines are approximately centered and evenly sized whiskers. Highest's median line is towards the top of the box, which means it is slightly left-skewed. 

```{r}
# Display the correlation matrix
cmw_corr = cor(cmw_merged[,-1])
cmw_corr
```

```{r}
# Correlation Matrix
library(corrplot)

corrplot(cmw_corr, order="hclust", title="Correlation Matrix of Coral Predictors",
         mar=c(0,0,1,0), cex.main=1)
```

Correlation Matrix interpretation:
In the above correlation matrix, apart from highest and lowest water levels, the intermediary predictors have a highly positive linear correlation between one another. Barometric pressure (baro) has almost no correlation with air temperature, degree heat week (dhw), and highest water level. Baro has a moderately negative correlation with all other predictors. Highest water level has almost no correlation with air temperature, bleaching alert area, and 7d max. Highest has a moderate to high positive correlation with all other variables. 

We use a threshold of 0.75 or -0.75 to determine highly correlated predictor variables. To reduce the issue of collinearity, we remove the field crw_baa_7d_max as it has a correlation value of 0.88036712. All other predictors live within the cutoff range.

```{r}
# Dropping highly correlated variables
coral_pretrain_v2 = subset(select(cmw_merged, -c(crw_baa_7d_max)))
```

At this point, the three data sets, coral, water level, and meteorological have been joined so that there are 6 measurements per day -- totaling 4386 observations, 16 predictor variables, and 1 response variable (crw_baa). All highly correlated predictors have been removed from the data set and all missing values have been imputed using the exponential moving average method. The data is now ready to be split for training.

```{r}
# Storing Final Set as csv
setwd("C:/Users/Rudy/Desktop/DataScience/deepseacorals/data/")
write.csv(coral_pretrain_v2, "coral_pretrain_v2.csv")
```

































