```{r}
setwd("C:/Users/Rudy/OneDrive/MADS/503_Applied Predictive Modeling/finalproject/data/")
# CSV Upload
met_21_22 = read.csv("met_21_22_9755371.csv")
met_22_23 = read.csv('met_22_23_9755371.csv')
wl_21_23 = read.csv('wl_21_23_9755371.csv')
pacIOOS_21_22 = read.csv('dhw_5km_6a81_de46_d4a4.csv')
pacIOOS_22_23 = read.csv('dhw_5km_db8d_5daf_80ed.csv')
```
```{r}
# Combine years for meteorological & coral tables
library('dplyr')
met21_23 = bind_rows(met_21_22, met_22_23)
coral21_23 = bind_rows(pacIOOS_21_22, pacIOOS_22_23)
```
```{r}
# Examine the meteorological tables
dim(met21_23)
colnames(met21_23)
str(met21_23)
```
```{r}
# Examine the water level table
dim(wl_21_23)
colnames(wl_21_23)
str(wl_21_23)
```
```{r}
# Examine the coral data
dim(coral21_23)
colnames(coral21_23)
str(coral21_23)
```
```{r}
# Convert coral table column names to lowercase
names(coral21_23) = tolower(names(coral21_23))
coral21_23
```
```{r}
# Convert coral datetime column to a standard date format. YYYY-MM-DD
## The coral timestamp field marks 6 data points at the beginning of each day.
### Extracting date from timestamp column
coral21_23$date = format(as.POSIXct(coral21_23$time), format="%Y-%m-%d")
```

There are now 3 tables each span 2021 through 2023 with a standard date format.

```{r}
# Count total missing values in each column of data frame
## Coral
sapply(coral21_23, function(x) sum(is.na(x)))
## Meteorological
sapply(met21_23, function(x) sum(is.na(x)))
## Water Level
sapply(wl_21_23, function(x) sum(is.na(x)))
```
```{r}

```

There are no missing values in any of the tables. 
---
All Coral _mask suffix columns contain only zeros and will be dropped from the data frame since we cannot confirm whether this an accurate or a missed reading based on the information provided from its source.

Coral columns to be dropped: crw_baa_mask, crw_dhw_mask, crw_hotspot_mask, and crw_ssanomaly_mask
---
The meteorological data contains 5 columns with not null empty values and shall be dropped.

Meteorological columns to be dropped: wind_speed, wind_direction, wind_gust, humidity, and visibility
---
The Water Level (WL) inf column contains no information and shall be dropped.

```{r}
# Dropping empty columns from each table
```


































